Robust Spectroscopic Inference with Imperfect Models

### 2014-09-11 tex/paper.tex
draft on interpolation errors roughly complete

We empirically determine the interpolation error by performing a series of ``drop-out'' interpolation tests, whereby we remove a spectrum from the grid, generate an interpolated spectrum from the nearest two grid points, and compare the difference between the two. For example, consider the {\sc Phoenix} grid used for the WASP-14 parameter estimation. In this parameter space, the grid is sampled every 100K in $T_\textrm{eff}$ and every 0.5 dex in $\log_{10} g$ and $\Z$. Therefore, for each spectrum in the synthetic grid, there are also three error spectra associated with it, one for each axis of stellar parameters. For example, consider a spectrum with stellar parameters $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex. A temperature ``error spectrum'' would be formed by subtracting this true spectrum from the spectrum formed by the average of the $T_\textrm{eff} = $ 6100K spectrum and the $T_\textrm{eff} = $ 5900K spectrum, with $\log_{10} g$ and $\Z$ held constant. Likewise, the $\log_{10} g$ and $\Z$ error spectra are created by performing the same operation over the $\log_{10} g$ and $\Z$ axes of the grid (see Figure~\ref{fig:interp_errors}).

These are done over a 200K range, and the actual range will only be 100K. Tests with other interpolation show that the interpolation error increases approximately linearly with interpolation distance. Therefore, when using the spectrum, we divide the total interpolation error spectrum by 2.

Generally, the interpolation error is small, typically no larger than 1\% across many lines. This test gives us confidence that it is possible to generate an interpolated spectra, and that inference on the stellar parameters may be drawn at a substantially finer level that the coarsity of the grid spacing itself, especially given the multitude of possible lines to fit. However, while the interpolation error is small, it is often the case that the error adds \emph{coherently}. When the signal to noise of the spectrum is particularly large, this coherent trend can manifest itself as a behavior that the spectral lines in an interpolated spectrum are less ``physical'' looking than those directly near a gridpoint, with minor or no interpolation. Therefore, it is necessary to quantify the degree of interpolation error, in order to give the proper likelihood to interpolated spectra.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_errors.pdf}
  \figcaption{Also showing the envelope for a two gridpoint interpolation. All errors measured relative to the value of the true spectrum.}
  \label{fig:interp_errors}
\end{center}
\end{figure}

Nearly all interpolation errors are below the spread in change for an interpolated value. Those few pixels for which the change in value is \emph{less} than the bracketed change are due to values for which the pixel value changes non-monotonically with parameters. In such a situation, because the synthetic grid is not Nyquist sampled, it is difficult to accurately predict what the true pixel value should be. In this case, the most conservative approach is to incorporate the large interpolation error budget, as we do.

@@

Because the interpolation error is covariant, i.e., the interpolation error spectrum is highly dependent on whether the pixel is on the continuum, or on a spectral line (and specifically where on the spectral line), we must model the covariance of the errors. We must multiply and then add in order to preserve the covariance. A simple sum would destroy the covariance and the effect.

For computational expediency and numerical stability, we smoothly taper the covariance to zero after a similar length scale as the global covariance kernel. In reality, this interpolation error matrix is dense: pixels separated by a wide spectral range may in fact exhibit a strong degree of correlation because they happen to fall on the same relative location of a spectral line. For example, all pixels at the continuum will tend to act similarly, regardless of where on the spectrum the pixels fall. Any pixel that falls on the blue side of a spectral line will tend to exhibit a similar error behaviour regardless of which spectral line, as long as that line was changing in a similar manner.

The elements of the interpolation covariance matrix $\mathsf{C}_{ij}^\textrm{int}$ are determined empirically from the drop-out spectra, and are a function of the degree of the interpolation. The variables are notated such that the last index changes most rapidly.

We could represent the average covariance matrix, for a specific axis $k$, completed from all of the grid points as

\begin{equation}
  \mathsf{C}_{ijk}^\textrm{int} = \sum_{l=1}^8  b_{kl} \lambda_{ikl} \lambda_{jkl}
\end{equation}
Then, the final combined matrix requires some weighting based upon interpolation distance in each of the three axes, $k$
\begin{equation}
  \mathsf{C}_{ij}^\textrm{int} = w_{ij}(r_0) \sum_{k=1}^3 a_k \mathsf{C}_{ijk}^\textrm{int}
\end{equation}
We also add in a Hann taper to ensure sparsity of the covariance matrix for computational efficiency, however note that the interpolation error covariance matrix is itself dense.

Here, $b_{kl}$ corresponds to the same weights that would be applied to each spectra in the interpolation case, for a specific axis $k$. The estimate of covariance is determined as a weighted average between the covariance matrices at all of the grid points.

Then, the final covariance

[Figure showing what the truncated covariance matrix looks like, with high values about lines].

>>>>

We empirically determine the interpolation error by performing a series of ``drop-out'' interpolation tests, whereby we remove a spectrum from the grid, generate an interpolated spectrum from the nearest two grid points, and compare the difference between the two. For example, consider the {\sc Phoenix} grid used for the WASP-14 parameter estimation. In this parameter space, the grid is sampled every 100K in $T_\textrm{eff}$ and every 0.5 dex in $\log_{10} g$ and $\Z$. Therefore, for each spectrum in the synthetic grid, there are also three error spectra associated with it, one for each axis of stellar parameters. For example, consider a spectrum with stellar parameters $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex. A temperature ``error spectrum'' would be formed by subtracting this true spectrum from the spectrum formed by the average of the $T_\textrm{eff} = $ 6100K spectrum and the $T_\textrm{eff} = $ 5900K spectrum, with $\log_{10} g$ and $\Z$ held constant. Likewise, the $\log_{10} g$ and $\Z$ error spectra are created by performing the same operation over the $\log_{10} g$ and $\Z$ axes of the grid (see Figure~\ref{fig:interp_errors}). Because the error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used (e.g., 50K), we reduce the amplitude of the error spectra by two. Drop-out tests at larger interpolation distances (e.g., 200K and 300K) show that the amplitude of the error spectrum grows approximately linearly with interpolation distance.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_errors.pdf}
  \figcaption{(\emph{top}) A small range of the ``true'' {\sc Phoenix} spectrum at $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex, shown with the drop-out interpolation spectra along each of the three parameter axes. These interpolation error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used in practice. Typically, the interpolated spectra closely match the ``true'' spectrum. (\emph{bottom}) The three ``error spectra'', or difference between the interpolated spectrum and the true spectrum, relative to the true spectrum value. The shaded region shows the total change in flux between the two spectra used for interpolation, relative to the true spectrum. Note that the error spectrum is generally much smaller than the typical change in flux. For the few points where this is not true (e.g., $\log_{10} g$ near 5161.5\AA) this implies a non-monotonic change in flux between grid points. Therefore, it is best to conservatively use the error spectrum, which protects against assumptions of monotonicity.}
  \label{fig:interp_errors}
\end{center}
\end{figure}

Generally, the interpolation error is small: typically no larger than 2\% of the flux and usually--but not always--much smaller than the total flux difference between spectra at adjacent grid points (see Figure~\ref{fig:interp_errors}). This test gives us confidence that it is possible to generate an interpolated spectra, and that inference on the stellar parameters may be drawn at a substantially finer level that the coarsity of the grid spacing itself, especially given the large number of possible lines to fit. However, while the interpolation error is small, because all spectral lines have roughly the same shape, it is often true that the error adds \emph{coherently}. When the signal to noise of the spectrum is particularly large, this coherent trend can contribute substantially by diminishing the likelihood of an interpolated spectrum, since the lines in the spectrum are less ``physical'' in shape than that of a spectrum directly on a gridpoint, which requires minor or no interpolation. Therefore in order to deliver the proper likelihood for interpolated spectra it is necessary to quantify the degree of interpolation error and add this into the covariance matrix.

@@

Because the interpolation error is highly dependent on whether the wavelength is on the continuum or on a spectral line, we must model the covariance of the errors. For example, all pixels at the continuum will tend to act similarly, regardless of where on the spectrum the pixels fall. Any pixel that falls on the blue side of a spectral line will tend to exhibit a similar error behaviour regardless of which spectral line, as long as that line was changing in a similar manner. The elements of the interpolation covariance matrix $\mathsf{C}_{ij}^\textrm{int}$ are determined empirically from the drop-out spectra, and are a function of the degree of the interpolation. Consider the manner in which a typical spectrum is tri-linearly interpolated from the synthetic grid (Figure~\ref{fig:interp_grid}). The interpolated spectrum is a weighted average of the neighboring 8 spectra, with weights $d_l$ based upon the distance from the grid points. To generate a composite covariance matrix which reflects the interpolation error present for the interpolated spectra, we must also combine the error spectra. Because the error is covariant, we must first multiply and then add the error spectra, in order to preserve the covariance. A simple sum of error spectra would destroy any information about the covariance. For a specific axis $k$ (e.g., $T_\textrm{eff}$, $\log_{10} g$, or $\Z$), we represent the average covariance matrix as a composite of all of the error spectra of this species
\begin{equation}
  \mathsf{C}_{ijk}^\textrm{int} = \sum_{l=1}^8  d_{kl} \lambda_{ikl} \lambda_{jkl}
\end{equation}

We combine the covariance matrices of each axis into a total interpolation covariance matrix by adding them together, with weights $g_k$ based upon the total interpolation distance in that axis. For example, if we were to interpolate a spectrum with $T_\textrm{eff} = 6120$K, then $g_{T_\textrm{eff}} = 0.4$. $g_{T_\textrm{eff}} = 0.0$ would signify no interpolation distance (0K, $T_\textrm{eff} = 6100$K) while $g_{T_\textrm{eff}} = 1.0$ would signify maximal interpolation distance (50K, $T_\textrm{eff} = 6150$K). Therefore, the final interpolation covariance matrix is
\begin{equation}
  \mathsf{C}_{ij}^\textrm{int} = w_{ij}(r_0) \sum_{k=1}^3 g_k \mathsf{C}_{ijk}^\textrm{int}
\end{equation}

For computational expediency and numerical stability, we smoothly taper the covariance to zero after a similar length scale as the global covariance kernel. In reality, the interpolation error matrix is dense: pixels separated by a wide spectral range may in fact exhibit a strong degree of correlation because they happen to fall on the same relative location of a spectral line. A typical covariance matrix is shown in Figure~\ref{fig:interp_matrix}.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_matrix.pdf}
  \figcaption{A typical interpolation error matrix for $T_\textrm{eff} = $ 6220K, $\log_{10} g=$4.2 dex, and $\Z = $ -0.4 dex.}
  \label{fig:interp_matrix}
\end{center}
\end{figure}
